---
title: "Super-short-intro to R and R Markdown"
author: "Giulia Grisot"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This is a super-short introduction to R and R Markdown.

## R and R Studio

R is a programming language and free software environment for statistical computing and graphics. It is widely used among statisticians and data miners for developing statistical software and data analysis. R is part of the GNU project and is available under the GNU General Public License. It is a command line language, but it can be used with a graphical user interface (GUI) such as RStudio. 

R Studio is an integrated development environment (IDE) for R. It includes a console, syntax-highlighting editor that supports direct code execution, as well as tools for plotting, history, debugging and workspace management.
While we rarely use R without R Studio, it is important to know that R Studio is not R.

R Studio makes it easier to use R, but also allows us to do more things, such as creating R Markdown documents, producing reports, presentations and websites.

You might be thinking 'well, I can do these things without R or R Studio'. This is true, but R and R Studio make it much easier to do these things, and they also make it easier to do them *well.*

How so? The answer is that R and R Studio are very flexible tools, as can be used for a wide range of tasks, from simple data manipulation to complex statistical analysis.
More importantly, R is a tool that allows us to do *reproducible research.* This is a very important aspect of research, as it allows us to share our work with others, and to make sure that our results are reliable.

For example, imagine you are working on a project and you have to do some data analysis. Without R and R Studio, you might do the following: you save the data in some excel files, you do some data manipulation, you run some statistical tests, and finally you create some plots. Ideally, you write down all the steps you have done in some Word document. When you save the output of your analysis, you write that down in another word document, or maybe your paper. Surely you need references for that, too, so yoiu might have another document with references in it, which you will need to copy paste in your paper. And you will need to do the same for all the numbers and data you have used.
*This is a lot of work, and it is very easy to make mistakes.* It is also very difficult to share your work with others, and to make sure that your results are reliable. 
*Crucially, if at some point you need to change something in your data or in your analysis, you will have to do all the work again*.

On the other hand, if you use R and R Studio, things are much smoother, less time-consuming, and more reliable.
You might still have your 'raw' data saved as an excel file, but you will do all the work in R.

You start by loading the data, then you do some data manipulation, then you run some statistical tests, and finally you create some plots. None of this is done manually, but it is all done with code, and none of this would risk modifying the original data.

You can write down all the steps you have done in an R script, or in a R Markdown document, where you also are actively proceeding with the analysis. There is no need to write down the output of your analysis, as it is already produced and saved in the document.

In fact, *you can write your paper directly in R Markdown*, and you can include the output of your analysis in the same document. You can also include the references you have used, and the code you have used. For the refrences, an easy way is to use a reference manager such as Zotero or Mendeley, and to save your library (ideally automatically) as a .bib file. You can then include the .bib file in your R Markdown document, and you can cite the references in the text directly. This is very useful, as it allows you to keep track of the references you have used, and to include them in your paper with a single click.

Finally, you can then share yiour raw data and yiour script with others, and they can see exactly what you have done, and they can reproduce your results. This is very important, as it allows others to verify your results, and to build on your work.

## Basic R commands

R is a command line language, which means that you can use it by typing commands in the console. For example, you can type `2+2` and press enter, and R will return `4`. You can also use R as a calculator, and you can do more complex operations, such as `2*2` or `2^2` (which is 2 to the power of 2).

Try it yourself! Type `2+2` in the console (below, after the > sign) and press enter. What do you get?

You can also use R to store values in variables, and to do more complex operations. For example, you can type `x <- 2` and press enter, and R will store the value 2 in the variable `x`. You can then use `x` in other operations, such as `x*2` or `x^2`. Or create vectors, which are sequences of values. For example, you can type `y <- c(1,2,3,4,5)` and press enter, and R will store the values 1,2,3,4,5 in the variable `y`. You can then use `y` in other operations, such as `y*2` or `y^2`.

You can also use R to create data frames, which are tables of values. For example, you can type `z <- data.frame(x=c(1,2,3,4,5), y=c(1,4,9,16,25))` and press enter, and R will store the values 1,2,3,4,5 in the variable `x` and the values 1,4,9,16,25 in the variable `y`. You can then use `z` in other operations, such as `z$x` or `z$y`.

Given we are using R Markdown, we can use the code chunks here to run R code. For example, we can use the following code chunk to create a data frame `z` and to print it.

```{r}

# Create a data frame
z <- data.frame(x=c(1,2,3,4,5), y=c(1,4,9,16,25))

# Print the data frame
print(z)

```

It is good practice to comment your code, as it makes it easier to understand what you have done. You can use the `#` symbol to comment your code. For example, you can see we wrote `# Create a data frame` before the line `z <- data.frame(x=c(1,2,3,4,5), y=c(1,4,9,16,25))`. This is a comment, and it is not executed by R.

## Useful R packages and functions

R has a lot of useful packages, which are collections of functions. One of the most useful ones is the `tidyverse`, which is a collection of packages for data manipulation and visualization. You can install the `tidyverse` package by typing `install.packages("tidyverse")` in the console and pressing enter. You can then load the package by typing `library(tidyverse)` in the console and pressing enter.

You can use the `tidyverse` package to do a lot of things, such as to manipulate data, to create plots, and to do statistical analysis. For example, you can use the `ggplot2` package to create plots, and you can use the `dplyr` package to manipulate data.

Importatnly, it allows you to write code that is easy to read and to understand, thakns to the 'piping' operator `%>%`. This is a very useful operator, as it allows you to write code that is easy to read and to understand. For example, you can use the `filter` function to filter rows in a data frame, and you can use the `mutate` function to create new columns in a data frame. You can then use the piping operator `%>%` to chain these operations together, and to create a new data frame.

For example, you can use the following code chunk to filter rows in the data frame `z` and to create a new data frame `z2`.

```{r}

# Load the tidyverse package
library(tidyverse)

# Filter rows in the data frame z and create a new data frame z2
# the filter function is used to filter rows in a data frame, in this case we are filtering rows where x is greater than 2
z2 <- z %>%
  filter(x>2)

# Print the new data frame
print(z2)


```

Pretty easy, right? You can see that we used the `filter` function to filter rows in the data frame `z`, and we used the piping operator `%>%` to chain this operation with the creation of a new data frame `z2`. We then printed the new data frame `z2`.

You can also use the `ggplot2` package to create plots. For example, you can use the following code chunk to create a scatter plot of the data frame `z`.

```{r}

# Create a scatter plot of the data frame z
# the ggplot function is used to create a plot, in this case we are creating a scatter plot of the data frame z with x on the x-axis and y on the y-axis

ggplot(z, aes(x=x, y=y)) +
  geom_point()


```

You can see that we used the `ggplot` function to create a plot, and we used the `aes` function to specify the x and y variables. We then used the `geom_point` function to create a scatter plot.

This is just a very brief introduction to R and R Studio, and to the `tidyverse` package. There is a lot more to learn, but I hope this was useful to you. If you have any questions, feel free to ask!

For the purpose of running our next script, you will be provided with all the scripts and data files you need. You will also be provided with a brief explanation of what the script does, and you will be asked to run the script and to answer some questions.

You are of course welcome to use the scripts and data files to do your own analysis, and learn from them.

Thanks!
Now let's clear our environmend and then move to the next script.

```{r}
rm(list=ls())
```





